One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2019-10-10 paper.tex
cutting out excursions waffle

\begin{abstract}
Point transect distance sampling in \texttt{inlabru}
\end{abstract}

@@

The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.  Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  Probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.

Over time interest grew in fully model-based approaches with a spatially explicit model for animal density that allow the use of non-randomized survey designs.  This also allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  
Model-based distance sampling has been implemented in a two-stage modelling framework.  In the first stage, detectability is estimated within each sampling unit.  In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.  Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.  Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.

A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}. Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.

One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.  The first efforts in this area used data augmentation approaches for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.  More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.

The reliance on MCMC for one-stage modelling presents a challenge for using spatially structured random effects, which can be computationally expensive.  \citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).  They considered line-transect distance sampling and simulateously estimated the detectability and spatial distribution of animals.
Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).  They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.  The observation process is formulated as a thinning of the log-Gaussian Cox process.  Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.

Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.  We replace the SPDE specification of the detection function with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.  Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.  Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.  We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.  The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.

To demonstrate the benefits of our approach we discuss issues regarding model evaluation and communication of results that are common across the field of species distribution modelling. We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.  A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much on summaries such as expectations and quantiles.  The one-stage Bayesian approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters. The spatial process and observation process are estimated jointly and so figures and summaries produced using posterior samples will naturally average over the uncertainty in the observation processes.  This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.

The article proceeds as follows.  First, we present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.  We also show how the point process model can be applied in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.  This is a common data format for distance sampling data.  Second, we describe how to specify and fit the joint model using an iterative approach to fitting models with components that are non-linear in their parameters - we call this the "iterated INLA" approach.  This allows us to avoid the SPDE approach to the detection function, instead using the parametric families common in the literature.  Third, we introduce our case-study of a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.  We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for the iterated INLA approach.  The paper includes example code to specify the model as well as full scripts for the analysis in the supplementary materials. We focus model evaluation and discussion on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.

@@

We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.  The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed effect covariates.

@@

The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.  In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.  If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.  The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.  Without such an assumption detectability and intensity are confounded.  The standard assumption of distance sampling is that the intensity is constant with respect to changes in $r(s)$.  Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.

A point transect distance sampling survey consists of a set of $K$ sampling units.  The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.  For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.  The thinning probability function $g(s)$ is defined relative to the positions of the sample units.  The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.  The probability of observing a point outside the surveyed region is zero.
Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.  For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$. The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation.  For simplicity we do not consider this here.  The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.

The thinne log-Gaussian Cox process likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
\begin{equation}
\label{lgcp-likelihood}
\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
\end{equation}
The integral component of the likelihood does not usually have analytical solutions.  Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.

@@

In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First, we specify the SPDE model as
\begin{lstlisting}
matern <- inla.spde2.pcmatern(mesh, 
                              prior.sigma = c(2, 0.01),    
                              prior.range = c(130, 0.01))
\end{lstlisting}
which uses the penalised complexity priors for the hyperparameters of the GMRF as introduced by \cite{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.

Second, we specify the thinning probability function.  The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
\begin{lstlisting}
thinning_function <- function(distance, lsig){
    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
    }
\end{lstlisting}
where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.  The dataset contains a column named \texttt{distance} that contains the observed distances.  The software will search for the appropriate columns using the names of the arguments specified in this function.  To allow the software to distinguish between names of covariates and parameters to be estimated we must also specify a model components object
\begin{lstlisting}
cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
\end{lstlisting} 
This object names all model components to be estimated.  Note that the \texttt{grf} component is a name we have chosen and not a function.  The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}. Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.

The model formula object is
\begin{lstlisting}
fml <- coordinates + distance ~ Intercept + grf +
	thinning_function(distance, lsig)
\end{lstlisting}
To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
\begin{lstlisting}
starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
W <- 60   # transect radius
distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
lgcp(components = cmp, 
     data = obs,
     samplers = samplers,
     domain = list(distance = distance_domain),
     formula = fml,
     options = list(result = starting_values, max.iter = 30))
\end{lstlisting}
where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.  Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.

@@

The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.  This shows a region of high intensity in the south and much lower intensity in the north.  The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.  There is clear preferential sampling and in the north estimated intensities are lower which will also inflate the posterior CV.  It is difficult to know whether the high CV in the north is primarily caused by lower sampling effort or lower estimated abundance, or some combination of the two.

\begin{figure}
	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
	\label{fig:intensity-mean-cv}
\end{figure}

The CV plot shows a measure of uncertainty relative to the mean intensity in each prediction grid cell.  However, often we are interested in the uncertainty around the mean intensity on the scale of the estimated intensity itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.  One approach attempt to communicate this uncertainty around the mean is to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).

\begin{figure}
	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.  All three plots use a common colour scale}
	\label{fig:intensity-quantiles}
\end{figure}

We believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.  The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.  We believe this misconception would be extremely common in particular for non-statistically trained audiences. The maps in \autoref{fig:intensity-quantiles} in fact only show \textit{summary statistics} of a posterior intensity which is specified (in our model) as a log-Gaussian random field.

This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.  Three such realizations are in \autoref{fig:intensity-realizations}.  There are some notable differences when compared to the mean and quantile summary maps.  Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.  The summary plots hide the spatial structure the model has been able to estimate from the data.  In general, the mean of a random field will contain less spatial structure than realisations of the field.

\begin{figure}
	\includegraphics[scale=0.35]{figures/intensity_realized.png}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}

The quantile plots are more concerning.  As mentioned above, the temptation is to interpret these as possible intensities that produced the observed data.  However, they in fact show the quantiles of the \textit{marginal} probability density for each prediction location. Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.  To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.  A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance. However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).  This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to interpretations (and possible management decisions) that are inconsistent with the very model that generated the maps.

As an alternative, we suggest using excursion sets and excursion functions \citep{bolin_excursion_2015}.  This is a recently developed approach to representing uncertainty in stochastic processes that specifically considers joint probability of events across multiple locations.  The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e. the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$. For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
\end{equation*}
It is important to note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.  This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.  Because $\lambda(s)$ is stochastic we can only ever say this is true with some probability and use something akin to a confidence level by specifying $\alpha$.  Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.  An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.

>>>>

\begin{abstract}
Distance sampling methods aim to estimate the size and distribution of animal populations.  Traditionally this has been achieved using a two-stage approach by using first-stage estimates of detectability as an offset in a second-stage generalised additive model.  One-stage approaches have typically been implemented using MCMC and data augmentation methods. Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability and spatial distribution.  We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.  We demonstrate the approach by applying the methods to point count survey data of Hawaiian \akepa{}. Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.    
\end{abstract}

@@

The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.  Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.

More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow the use of non-randomized survey designs.  These methods also allow density to be associated with spatially-indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.

Model-based distance sampling has been implemented in a two-stage modelling framework.  In the first stage, detectability is estimated within each sampling unit.  In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.  Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.  Negative-binomial and Tweedie distributions are common choices to model the distribution of counts.  This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review and software package \texttt{dsm} to fit the models.

A key concern with the two-stage approach is the propagation of the uncertainty from the detectability estimates to the second-stage spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}. Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.

One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.  The first efforts in this area used data augmentation approaches to account for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.  More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.

The reliance on MCMC for one-stage modelling presents a challenge for using spatially-structured random effects, which can be computationally expensive.  \citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).  Applying their method to line-transect distance sampling, they simulateously estimated the detectability and spatial distribution of animals.
Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).  They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.  The observation process is formulated as a thinning of the log-Gaussian Cox process.  Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.

Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.  Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.

Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.  We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.  It is common for distance sampling surveys to only record the distance to the observer and the observer location.  We show how point process methods can handle this type of data even though the exact location of the point is not known.

We replace the SPDE specification of the detection function in \cite{yuan_point_2017} with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.  This results in components of the additive predictor that are non-linear in their parameters.  To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.  We name this approach \textit{iterated INLA}.

To the best of our knowledge we also demonstrate the first use of a sparse GMRF spatial random effect with point transect distance sampling data.  The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}. The flexibility of the mesh allows a higher-resolution discretisation of space where complexity requires it but also lower resolutions in areas where this is not required. As collecting high-resolution data and incorporating complex spatial domains (broadly defined here as domains where regular lattice discretisations of space lead to a computational bottleneck) becomes more common we believe this approximate-Bayesian GMRF approach will become a valuable addition to the spatial ecology toolbox.

% DRASTICALLY REWORK NEXT TWO PARAGRAPHS

We apply the methods to point transect survey data of Hawaiian \akepa{}, an endangered bird that is the focus of targeted conservation efforts.  We simultaneously estimate the detectability and spatial distribution of \akepa{} in a region of Hawaii island.  Model evaluation focuses on uncertainty in model outputs that are based on sampling from the joint posterior of all model parameters.  This provides a natural way to produce useful outputs from the analysis that average over detection model uncertainty.  The posterior intensity field is investigated in detail and its uncertainty visualised using excursion sets and excursion functions \citep{bolin_excursion_2015, bolin_calculating_2018}, approaches which deal with the problem of multiple hypothesis testing when mapping prediction quantiles.

The article proceeds as follows:
\begin{enumerate}[(i)]
\item We present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.  We also show how the point process model can be applied in the case of incomplete location information. 
\item We describe how the model can be specified in a modified Poisson log-likelihood form and fitted using the iterated INLA approach.
\item A case-study using point transect distance sampling survey of the Hawaiian \akepa{}.  We give example code showing how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for iterated INLA.
\end{enumerate}
Full scripts for the analysis can be found in the supplementary materials.

@@

We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.  The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed-effect covariates.

@@

The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.  In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.  If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.  The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.  Without such an assumption detectability and intensity are confounded.  The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.  Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.

A point transect distance sampling survey consists of a set of $K$ sampling units.  The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.  For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.  The thinning probability function $g(s)$ is defined relative to the positions of the sample units.  The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.  The probability of observing a point outside the surveyed region is zero.
Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.  For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$. The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation but, for simplicity, we do not consider this here.  The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.

The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
\begin{equation}
\label{lgcp-likelihood}
\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
\end{equation}
The integral component of the likelihood does not usually have analytical solutions.  Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.

@@

In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate how to specify the model.  First, we specify the SPDE model component as
\begin{lstlisting}
matern <- inla.spde2.pcmatern(mesh, 
                              prior.sigma = c(2, 0.01),    
                              prior.range = c(130, 0.01))
\end{lstlisting}
which uses penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.  The value of 130 was selected based on the minimum distance between sampling locations.

Second, we specify the thinning probability function model component.  The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
\begin{lstlisting}
thinning_function <- function(distance, lsig){
    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
    }
\end{lstlisting}
where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.  The dataset contains a column named \texttt{distance} that contains the observed distances.  The software will search for the appropriate columns using the names of the arguments specified in this function.  To allow the software to distinguish between named covariates and parameters - which both appear as arguments of the function to be estimated - we must also specify a model components object
\begin{lstlisting}
cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
\end{lstlisting} 
This object names all model components to be estimated.  Note that the \texttt{grf} component is a name we have chosen and not a function.  The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}. Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.

We can now specify the predictor formula using the components defined above.  
\begin{lstlisting}
fml <- coordinates + distance ~ Intercept + grf +
	thinning_function(distance, lsig)
\end{lstlisting}
To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
\begin{lstlisting}
starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
W <- 60   # transect radius
distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
lgcp(components = cmp, 
     data = obs,
     samplers = samplers,
     domain = list(distance = distance_domain),
     formula = fml,
     options = list(result = starting_values, max.iter = 30))
\end{lstlisting}
where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.  We also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.  Full scripts for the analysis are available in the supplementary materials.

@@

The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.  This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary materials).  The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.  There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
\begin{figure}
	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
	\label{fig:intensity-mean-cv}
\end{figure}

The fact that CV values are influenced both by sampling effort and the estimated intensity means other methods to visualise uncertainty in the posterior intensity field are also often considered. A common approach is to map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
\begin{figure}
	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.  All three plots use a common colour scale}
	\label{fig:intensity-quantiles}
\end{figure}
Although commonly used, we believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.  The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.  Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.  To demonstrate the fallacy of this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.  A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance. However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).

As an alternative to the quantile maps we use excursion sets and excursion functions \citep{bolin_excursion_2015}.  This is a recent addition to the literature so we take some time here to define and elaborate on the concepts used.  This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.  The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e. the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$. For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
% Somehow keep this information somewhere in the paper:
%This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.  Three such realizations are in \autoref{fig:intensity-realizations}.  There are some notable differences when compared to the mean and quantile summary maps.  Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.  In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.  The summary plots hide the spatial structure the model has been able to estimate from the data.  In general, the mean of a random field will contain less spatial structure than realisations of the field.
%
%\begin{figure}
%	\includegraphics[scale=0.35]{figures/intensity_realized.png}
%	\caption{Three realizations of the posterior intensity field}
%	\label{fig:intensity-realizations}
%\end{figure} 
\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
\end{equation*}
Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.  As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.  Negative excursion sets can be similarly defined.  Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.  An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
